Microxanox: an R package for simulating an aquatic \(MICR\)obial ecosystem that can occupy \(OX\)ic or \(ANOX\)ic states.

### 2023-05-15 inst/manuscript/manuscript_microxanox.tex
WORKING ON TEX ARTICLE!

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Microxanox: an R package for simulating an aquatic MICRobial ecosystem that can occupy OXic or ANOXic states.},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}

@@

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

@@

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
\usepackage{lineno}

@@

%\usepackage{lineno}
%\linenumbers
%\usepackage{setspace}
%\doublespacing
%\usepackage{fancyhdr}
%\fancyhead{}
%\fancyfoot{}
%\fancyhead[CO,CE]{Microxanox R package}
%\pagestyle{fancy}
%

@@

  \title{Microxanox: an R package for simulating an aquatic
\(MICR\)obial ecosystem that can occupy \(OX\)ic or \(ANOX\)ic states.}
    \author[University of Zürich]{Rainer M Krug%
  %
  \fnref{Corresponding Author}}
   \ead{Rainer.Krug@uzh.ch; Rainer@krugs.de} 
    \author[University of Zürich]{Owen L. Petchey}
   \ead{Owen.Petchey@ieu.uzh.ch} 
      \affiliation[University of Zürich]{Department of Evolutionary
Biology and Environmental Studies, Winterthurerstrasse 190, 8057 Zurich}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{Corresponding Author}
    \fntext[2]{Equal contribution}

  \begin{abstract}
  \(Microxanox\) is an R package to simulate a three functional group
  ecosystem (cyanobacteria, phototrophic sulfur bacteria, and
  sulfate-reducing bacteria) with four chemical substrates (phosphorus,
  oxygen, reduced sulfur, and oxidized sulfur) using a set of ordinary
  differential equations. Simulations can be run individually or over a
  parameter range, to find stable states. The model can be implemented
  with different numbers of species per functional group. The package is
  constructed in such a way that the results contain the input parameter
  used, so that a saved results can be loaded again and the simulation
  be repeated. Furthermore, the package framework and code should serve
  as a useful starting point for making simulation models of other types
  of ecosystem.
  \end{abstract}
    \begin{keyword}
    reproducibility \sep regime shift \sep stable state \sep 
    ordinary differential equations
  \end{keyword}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0483}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4138}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5379}}@{}}
\toprule
\begin{minipage}[b]{\linewidth}\centering
\textbf{Nr.}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Code metadata description}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Please fill in this column}
\end{minipage} \\
\midrule
\endhead
C1 & Current code version & v0.9.0 \\
C2 & Permanent link to code/repository used for this code version &
\url{https://github.com/UZH-PEG/microxanox} \\
& & \href{https://zenodo.org/record/6624125}{10.5281/zenodo.6624125} \\
C3 & Code Ocean compute capsule & \\
C4 & Legal Code License & MIT \\
C5 & Code versioning system used & git \\
C6 & Software code languages, tools, and services used &
\href{https://cran.r-project.org/index.html}{R} \\
C7 & Compilation requirements, operating environments &
\href{https://cran.r-project.org/index.html}{R (\textgreater= 4.1.0)} \\
& & magrittr \\
& & tibble \\
& & ggplot2 \\
& & patchwork \\
& & grDevices \\
& & stats \\
& & mgcv \\
& & deSolve \\
& & dplyr \\
& & tidyr \\
& & stringr \\
& & multidplyr \\
C8 & If available Link to developer documentation/manual &
\href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{User
Guide} \\
C9 & Support email for questions & Rainer.Krug@uzh.ch;
Owen.Petchey@ieu.uzh \\
\bottomrule
\end{longtable}

@@

Mathematical models play a key role in the development of understanding
about how ecosystems work and how they respond to environmental changes
\citep{Binzer2016a, Chaparro-Pedraza2021, Vasseur2005}. They are also
important for developing hypotheses to test in empirical studies. One
area of ecology in which models have played a influential role is how
ecosystems respond to gradual change in an environmental driver
\citep{Scheffer2001}. An environmental driver is a environmental
condition that affects an ecosystem, but is assumed to not be affected
by the ecosystem, such as the rate of nutrient input into a lake.

@@

An example where a gradual change of an environmental variable causes an
abrupt change of the system is the switch from an aerobic (oxygen is
available for metabolism) to anaerobic (oxygen generally unavailable)
state in a microbial ecosystem. This system has been investigated by
\citet{Bush2017} in a simulation study of a mathematical model. Three
types of microbes occur in the model: cyanobacteria (CB) dominating the
oxic state, and two types of sulfur bacteria that dominate the anoxic
state (sulfate reducing bacteria (SB) and phototrophic sulfur bacteria
(PB)). The model shows that gradual change in the rate at which oxygen
could diffuse into the ecosystem (termed the oxygen diffusitivity) could
cause catastrophic changes in the ecosystem state that would be
difficult to reverse.

One feature of the study by \citet{Bush2017} was limited biodiversity.
Specifically, there was no biodiversity within each of the three types
of bacteria. This leaves open the question of if and how biodiversity
within these types (i.e.~functional groups) of bacteria affects the
ecosystem response to environmental change. This limitation is not
specific to the study of \citet{Bush2017}. There are few if any studies
of the effects of biodiversity on abrupt transitions between ecosystem
states.

We decided to fill this research gap by making a simulation study of how
within functional group biodiversity affects ecosystem responses to
environmental change \citet{Limberger2022}, and to base our work on the
work and model of \citet{Bush2017}. It was with this goal in mind that
we developed the \emph{microxanox} package \citep{Krug2022}. The first
stage of development was to write code from scratch (as there was no
available code to start from) and to confirm that this new
implementation would reproduce the previously published results. The
resulting reproduction is available as one of the package vignettes:
\href{https://uzh-peg.r-universe.dev/articles/microxanox/partial-reproduction-Bushetal2017.html}{vignette
Partial reproduction of Bush et al}.

The second stage was to add functionality that would be necessary to
answer our research question. Most importantly, we made it possible to
have multiple species of bacteria within each of the three functional
groups, for the multiple species to differ in their characteristics, and
to vary the number of species and amount of variability among them. We
also added functionality that allowed: temporally varying environmental
conditions, addition of random noise to state variables, and
immigration. In addition to the model itself, the package includes some
functions to analyse the results as well as to visualize the results to
provide a starting point for customized visualizations based on own
requirements. The basic and additional functionality is described in the
package
\href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{User
Guide}.

@@

The \emph{microxanox} package is for simulating a three functional group
system (\emph{CB}: cyanobacteria, \emph{PB}: phototrophic sulfur
bacteria, and \emph{SB}: sulfate-reducing bacteria) with four chemical
substrates (\emph{P}: phosphorus, \emph{O}: oxygen, \emph{SR}: reduced
sulfur, \emph{SO}: oxidized sulfur). It includes feedback between
organisms and biogeochemical processes and is based on \citet{Bush2017}
(See \citet{Bush2017} for a detailed discussion of the model). At the
core of the simulations is a set of ordinary differential equations
(specified in the function \texttt{bushplus\_dynamic\_model()}, though
this function need not be directly called). There are functions for
running individual simulations and for running a set of simulations
across, for example, a range of environmental conditions.

To make the simulation run with multiple species per functional group,
we expressed different species characteristics in the elements of
vectors and matrices. We also coded the ordinary differential equations
to include the vectors and matrices, and to use matrix mathematics. In
this way, we made it possible to run simulations with different numbers
of species without having to change the underlying code.

The package functions and code have modular structure so that new
functionality can be easily added. E.g. temporally defined events of any
type could be specified. Further, all parameter values required to run a
simulation are stored in one object. Lastly, the general structure of
the code should make it straightforward to adapt the model to other
similar systems (described in more detail in the Impact section).

@@

The framework used when writing this package aimed to maximise
simplicity for the user, and to make it straightforward to reproduce
results (see the supplement \citep{Petchey2022} to \citet{Limberger2022}
for an example of how this is used). As such, all the parameters needed
to run a simulation or find a stable state (i.e.~the final state of the
ecosystem) are contained in a single object (which can easily be created
using included functions). This parameter object is given to a function
that runs the simulations and returns the results. The returned results
object is identical to the parameter object but with an additional slot
named \texttt{results}, which contains the results of the run. Thus the
returned results object contains the simulation conditions (parameters)
as well as the results, and can be used to run the simulation again.
This promotes reproducibility and makes incremental changes of
individual parameters and re-running the simulations straightforward.

@@

A simulation is run using the \texttt{run\_simulation()} function. In
this function, the ODEs are solved using the function \texttt{ode()} in
the package \emph{deSolve} package \citep{Soetaert2010}. The
\texttt{run\_simulation()} function needs only one argument - an object
as created by the function \texttt{new\_runsim\_parameter()}. The
parameter object returned by \texttt{new\_runsim\_parameter()} contains
among other things the \texttt{strain\_parameter} object, which can be
created by the function \texttt{new\_strain\_parameter()}. For a
detailed description of the parameter objects, their meaning and how
they are created and have values set and changed please see the
\emph{User Guide} which accompanies the package or is available at
\href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{User
Guide}.

After the parameter object has been defined, it can be used in the
\texttt{run\_simulation()} function. The function returns an object
which is identical to the parameter object, except of an additional slot
containing the results. This design produces a fully reproducible object
as it can be used as a parameter object to be fed back into the
\texttt{run\_simulation()} function to run the simulation again from the
parameter used to generate the results before.

@@

\renewcommand\refname{References}
\bibliography{bibliography.bib}

>>>>

 \documentclass[preprint,12pt, a4paper]{elsarticle}

%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
% \usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}

\usepackage[unicode=true]{hyperref}

\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Microxanox: an R package for simulating an aquatic MICRobial ecosystem that can occupy OXic or ANOXic states.},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}

@@

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother

@@

@@

@@

  \title{Microxanox: an R package for simulating an aquatic
\(MICR\)obial ecosystem that can occupy \(OX\)ic or \(ANOX\)ic states.}
    \author[University of Zürich]{Rainer M Krug%
  %
  \fnref{Corresponding Author}}
   \ead{Rainer.Krug@uzh.ch; Rainer@krugs.de} 
    \author[University of Zürich]{Owen L. Petchey%
  %
  }
   \ead{Owen.Petchey@ieu.uzh.ch} 
      \affiliation[University of Zürich]{Department of Evolutionary
Biology and Environmental Studies, Winterthurerstrasse 190, 8057 Zurich}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{Corresponding Author}
    \fntext[2]{Equal contribution}

  \begin{abstract}
  \(Microxanox\) is an R package to simulate a three functional group
  ecosystem (cyanobacteria, phototrophic sulfur bacteria, and
  sulfate-reducing bacteria) with four chemical substrates (phosphorus,
  oxygen, reduced sulfur, and oxidized sulfur) using a set of ordinary
  differential equations. Simulations can be run individually or over a
  parameter range, to find stable states. The model can be implemented
  with different numbers of species per functional group. The package is
  constructed in such a way that the results contain the input parameter
  used, so that a saved results can be loaded again and the simulation
  be repeated. Furthermore, the package framework and code should serve
  as a useful starting point for making simulation models of other types
  of ecosystems.
  \end{abstract}
    \begin{keyword}
    reproducibility \sep regime shift \sep stable state \sep 
    ordinary differential equations
  \end{keyword}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0483}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4138}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5379}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
\textbf{Nr.}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Code metadata description}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Please fill in this column}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
C1 & Current code version & v0.9.0 \\
C2 & Permanent links to code/repository used for this code version &
\url{https://github.com/UZH-PEG/microxanox} \\
& & \href{https://zenodo.org/record/6624125}{10.5281/zenodo.6624125} \\
C3 & Code Ocean compute capsule & \\
C4 & Legal Code License & MIT \\
C5 & Code versioning system used & git \\
C6 & Software code languages, tools, and services used &
\href{https://cran.r-project.org/index.html}{R} \\
C7 & Compilation requirements, operating environments &
\href{https://cran.r-project.org/index.html}{R (\textgreater= 4.1.0)} \\
& & magrittr \\
& & tibble \\
& & ggplot2 \\
& & patchwork \\
& & grDevices \\
& & stats \\
& & mgcv \\
& & deSolve \\
& & dplyr \\
& & tidyr \\
& & stringr \\
& & multidplyr \\
C8 & If available Link to developer documentation/manual &
\href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{User
Guide} \\
C9 & Support email for questions & Rainer.Krug@uzh.ch;
Owen.Petchey@ieu.uzh \\
\end{longtable}

@@

Mathematical models play a key role in the development of ecosystem
models and the understanding about how ecosystems work and respond to
environmental changes
\citep{Binzer2016a, Chaparro-Pedraza2021, Vasseur2005}. They are also
important for developing hypotheses to test in empirical studies. One
area of ecology in which models have been influential is the
understanding of their response to a gradual change in an environmental
driver \citep{Scheffer2001}. An environmental driver is an environmental
condition that affects an ecosystem. Still, it is assumed not to be
affected by the ecosystem, such as the rate of nutrient input into a
lake.

@@

An example where a gradual change of an environmental variable causes an
abrupt change in the system is the switch from an aerobic (oxygen is
available for metabolism) to an anaerobic (oxygen generally unavailable)
state in a microbial ecosystem. This system has been investigated by
\citet{Bush2017} in a simulation study of a mathematical model. Three
types of microbes occur in the model: cyanobacteria (CB) dominating the
oxic state, and two types of sulfur bacteria that dominate the anoxic
state (sulfate reducing bacteria (SB) and phototrophic sulfur bacteria
(PB)). The model shows that gradual change in the rate at which oxygen
could diffuse into the ecosystem (termed oxygen diffusitivity) could
cause catastrophic changes in the ecosystem state that would be
difficult to reverse.

One feature of the study by \citet{Bush2017} was limited biodiversity.
Specifically, there was no biodiversity within each of the three types
of bacteria. This leaves the question of if and how bacteria's
biodiversity within these types (i.e., functional groups) affects the
ecosystem response to environmental change. This limitation is not
specific to the study of \citet{Bush2017}. There are few if any studies
of the effects of biodiversity on abrupt transitions between ecosystem
states.

We decided to fill this research gap by making a simulation study of how
within functional group biodiversity affects ecosystem responses to
environmental change \citet{Limberger2022}, and to base our work on the
work and model of \citet{Bush2017}. With this goal in mind, we developed
the microxanox package \citep{Krug2022}. The first stage of development
was to write code from scratch (as there was no available code to start
from) and to confirm that this new implementation would reproduce the
previously published results. The resulting reproduction is available as
one of the package vignettes:
\href{https://uzh-peg.r-universe.dev/articles/microxanox/partial-reproduction-Bushetal2017.html}{vignette
Partial reproduction of Bush et al}.

The second stage was to add functionality that would be necessary to
answer our research question. Most importantly, we made it possible to
have multiple species of bacteria within each of the three functional
groups, for the multiple species to differ in their characteristics, and
to vary the number of species and amount of variability among them. We
also added functionality that allowed: temporally varying environmental
conditions, the addition of random noise to state variables, and
immigration. In addition to the model itself, the package includes some
functions to analyse the results and visualize these to provide a
starting point for customized visualizations based on own requirements.
The basic and additional functionality is described in the package
\href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{User
Guide}.

@@

The \emph{microxanox} package is for simulating a three functional group
system (\emph{CB}: cyanobacteria, \emph{PB}: phototrophic sulfur
bacteria, and \emph{SB}: sulfate-reducing bacteria) with four chemical
substrates (\emph{P}: phosphorus, \emph{O}: oxygen, \emph{SR}: reduced
sulfur, \emph{SO}: oxidized sulfur). It includes feedback between
organisms and biogeochemical processes and is based on \citet{Bush2017}
(See \citet{Bush2017} for a detailed discussion of the model). At the
core of the simulations is a set of ordinary differential equations
(specified in the function \texttt{bushplus\_dynamic\_model()}, though
this function need not be directly called). There are functions for
running individual simulations and a set of simulations across, for
example, a range of environmental conditions.

To make the simulation run with multiple species per functional group,
we expressed different species characteristics in the elements of
vectors and matrices. We also coded the ordinary differential equations
to include the vectors and matrices and used matrix mathematics. In this
way, we made it possible to run simulations with different numbers of
species without having to change the underlying code.

The package functions and code have a modular structure so that new
functionality can be easily added. E.g., temporally defined events of
any type could be specified. Further, all parameter values required to
run a simulation are stored in one object. Lastly, the general structure
of the code should make it straightforward to adapt the model to other
similar systems (described in more detail in the Impact section).

@@

The framework used when writing this package aimed to maximise
simplicity for the user and to make it straightforward to reproduce
results (see the supplement \citep{Petchey2022} to \citet{Limberger2022}
for an example of how this is used). As such, all the parameters needed
to run a simulation or find a stable state (i.e.~the final state of the
ecosystem) are contained in a single object (which can easily be created
using included functions). This parameter object is given to a function
that runs the simulations and returns the results. The returned results
object is identical to the parameter object but with an additional slot
named \texttt{results}, which contains the simulation results. Thus the
returned results object contains the simulation conditions (parameters)
as well as the results, and can be used to run the simulation again.
This promotes reproducibility and makes incremental changes of
individual parameters with a consecutive re-running of the simulations
straightforward.

@@

A simulation is run using the \texttt{run\_simulation()} function. In
this function, the ODEs are solved using the function \texttt{ode()} in
the package \emph{deSolve} package \citep{Soetaert2010}. The
\texttt{run\_simulation()} function needs only one argument - an object
as created by the function \texttt{new\_runsim\_parameter()}. The
parameter object returned by \texttt{new\_runsim\_parameter()} contains,
among other things, the \texttt{strain\_parameter} object, which can be
created by the function \texttt{new\_strain\_parameter()}. For a
detailed description of the parameter objects, their meaning and how
they are created and have values set and changed, please see the
\emph{User Guide} which accompanies the package or is available at
\href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{User
Guide}.

After the parameter object has been defined, it can be used in the
\texttt{run\_simulation()} function. The function returns an object
identical to the parameter object, except of an additional slot
containing the results. This design produces a fully reproducible object
as it can be used as a parameter object to be fed back into the
\texttt{run\_simulation()} function to rerun the simulation again from
the parameter used to generate the results before.

@@

\renewcommand\refname{References}
\bibliography{bibliography}
\bibliographystyle{elsarticle-num-names.bst}
